---
title: "Variation within blocks for individual plant variables"
output:
  html_document:
    code_folding: hide
  html_notebook: default
  pdf_document: default
fontsize: 14pt
---
<style type="text/css">

body{ /* Normal  */
      font-size: 16px; }
td {  /* Table  */
  font-size: 10px; }
h1.title {
  font-size: 42px;
  color: DarkRed; }
h1 { /* Header 1 */
  font-size: 32px;
  color: DarkRed; }
h2 { /* Header 2 */
    font-size: 25px;
  color: DarkRed; }
h3 { /* Header 3 */
  font-size: 22px;
  font-family: "Times New Roman", Times, serif;
  color: DarkRed; }
code.r{ /* Code block */
    font-size: 12px; }
pre { /* Code block - determines code spacing between lines */
    font-size: 14px; }
</style>

&nbsp;

Here we present analysis of the "individual_traits" sheet of "plant_traits_DPE" table. The table represents data for the "Height, cm", "Length, cm", "Stem diameter, mm", "Weight of all wet leaves, g", "Weight of all dry leaves, g", "Area of all leaves, cm2", "Specific leaf area, cm2/g" variables. Each variable representad by 9 observations (3 observations for 3 species) within each of 3 plots within each of 10 blocks.

We want to test if there are any significant differences between the plots in each block.

```{r, message=FALSE, warning=FALSE}
library(xlsx)
library(ggplot2)
library(ggpubr)
library(plyr)
library(lme4)
source("./functions.R")
```
```{r, message=FALSE, warning=FALSE}
path_ins <- "/home/lena/Tundra_exp/plant_traits_DPE.xlsx"
ins <- read.xlsx(path_ins, 2)

clmn = colnames(ins) %in% c("block", "plot",  "species", "height_cm", "length_cm", "stem_diameter_mm", 
                            "all_leaves_wet_g", "all_leaves_dry_g",  "all_leaves_cm2", "SLA_cm2_g")
ins = ins[!is.na(ins$block),clmn]
rm(clmn, path_ins)

cols = c(4:ncol(ins))
ins[,cols] = apply(ins[,cols], 2, function(x) as.numeric(as.character(x)))
ins[,1] = factor(ins[,1],levels(ins[,1])[c(1,3:10,2)])
na_obs = sapply(c(1:nrow(ins)), function(x) all(is.na(ins[x,c(4:ncol(ins))])))
ins = ins[!na_obs,]

levels(ins$plot) = substr(levels(ins$plot), 2, 4)

ynames = colnames(ins)[4:ncol(ins)]
varnames = c("Height, cm", "Length, cm", "Stem diameter, mm", "Weight of all wet leaves, g", "Weight of all dry leaves, g",  
             "Area of all leaves, cm2", "Specific leaf area, cm2/g")
```


# Statistical tests
&nbsp;

In order to test differences within blocks, we perform ANOVA for blocks that meet ANOVA's assumptions and Kruskal-Wallis test for others
```{r}

test = sapply(c(4:ncol(ins)), function(cl) {
  
  block_range = c(1:10)
  
  # ANOVA test
  anpv = sapply(block_range , function(x) {
    indexes_within_block = (ins$block == levels(ins$block)[x])
    formula_modl = as.formula(paste0(colnames(ins)[cl]," ~ plot"))
    modl = aov(formula_modl, data = ins[ indexes_within_block ,])
    res = anova(modl)$`Pr(>F)`                                         # results of ANOVA
    res[2] = shapiro.test(residuals(modl))$p.value                     # normality of residuals
    res[3] = bartlett.test(ins[ indexes_within_block ,cl], 
                           ins$plot[ indexes_within_block ])$p.value    # homogeneity of variances
    
    tk = TukeyHSD(modl)$plot
    res[4] = 0
    if (sum(tk[,4] < 0.05) > 0)
      res[4] = max(abs(tk[tk[,4] < 0.05,1]))
    return(res)
  })

  non_anova_blocks = which((anpv[2, ] < 0.05) | (anpv[3, ] < 0.05))    # block's numbers that do not meet
                                                                       # ANOVA's assumptions

  if (length(non_anova_blocks) !=0) {
    # Kruskal-Wallis for non-normal residuals
    krpv = sapply( block_range[non_anova_blocks] , function(x) {
      indexes_within_block = (ins$block == levels(ins$block)[x])
      res = kruskal.test(ins[ indexes_within_block , cl], ins$plot[ indexes_within_block ])$p.value
      pw = pairwise.wilcox.test(ins[indexes_within_block,cl], ins$plot[indexes_within_block  ], exact = F)
      pw = as.vector(pw$p.value)[c(1,2,4)]
      if (sum(pw < 0.05) > 0) {
        means = sapply(c(2:0), function(y) mean(ins[ins$plot == levels(ins$plot)[x*3 - y], cl],na.rm = T))
        res[2] = max(abs(apply(combn(means,2), 2, diff))[pw < 0.05])
      } else res[2] = 0
      
      return(res)
    })
    
  if (length(non_anova_blocks) == 10)  {
    diff_blocks = block_range[krpv[1,] <0.05]
    max_diff = max(krpv[2,])
  } else  {
    # Result of ANOVA for normal residuals and Kruskal-Wallis for non-normal residuals
    diff_blocks = c(setdiff(block_range[anpv[1,] <0.05], block_range[non_anova_blocks]), (block_range[non_anova_blocks])[which(krpv[1,] <0.05)])
    max_diff = max(c(anpv[4,-non_anova_blocks], krpv[2,]))
    }
  
  } else {
      diff_blocks = block_range[which(anpv[1,] <0.05)]                    # Result of ANOVA
      max_diff = max(anpv[4,]) 
      }
  
  cat("There are significant differences within ", length(diff_blocks), " blocks of '",
               varnames[cl-3], "' variable.\n", sep = "")
  if (length(diff_blocks) > 0) {
    cat("The differences are in blocks number", diff_blocks, "\n")
    
    cat("The maximum significant difference between means within these blocks is", round(max_diff,2), "g.\n")
  }
  
  cat("\n")
  return(round(max_diff,2))

})
```
Note, that all species are analysed together.

# Graphical representation of differences within blocks and between blocks
&nbsp;
 
On the figure each dot represents the mean of the variable within plot, vertical lines represent confidance intervals. The plot shows for each variable it's variation within each block. Different colors indicate different blocks.

```{r, fig.height=15, fig.width=10, message=FALSE, warning=FALSE}
plot_list <- lapply(c(1:length(ynames)), function(n) plot_diff(ins, ynames[n], varnames[n], two_plots = F))
cowplot::plot_grid(plotlist=plot_list, ncol=1)
```



# Conclusion
&nbsp;
There are no significant differences for individual plant variables within blocks, except for 'Weight of all wet leaves' and 'Weight of all dry leaves' variables. The mean differences for these variables are:
```{r}
concl = sapply(c(7:8), function(cl) {
  vname = substr(varnames[cl - 3], 1, nchar(varnames[cl - 3]) - 3)
  cat("* for '", vname, "' variable (", min(ins[,cl], na.rm = T), " — ", max(ins[,cl], na.rm = T), " g) up to ",test[cl-3], " g.\n", sep = "")
})
```

&nbsp;

