---
title: "Plant traits"
output: html_notebook
---

```{r}
library(xlsx)
library(ggplot2)
library(ggpubr)
source("./functions.R")
```

# Canoppy traits

```{r Reading the canopy sheet of plant_traits table, message=FALSE, warning=FALSE}
can_path <- "/home/lena/Tundra_exp/plant_traits_DPE.xlsx"
can <- read.xlsx(can_path, 3)

clmn = colnames(can) %in% c("block", "plot", "shrub_height_cm", "forb_height_cm", "graminoid_height_cm", "moss_layer_thickness_cm", "litter_layer_thickness_cm")
can = can[!is.na(can$block),clmn]
cols = c(3:7)
can[,cols] = apply(can[,cols], 2, function(x) as.numeric(as.character(x)))
can[,1] = factor(can[,1],levels(can[,1])[c(1,3:10,2)])

rm(clmn, can_path)
```

Number of pairs within plot that have significant differences according to Mann-Witney test
```{r}
n_samp = 10
cl =  6
tpv1 = sapply(c(0:9), function(x) wilcox.test(can[(1 + n_samp*3*x):(n_samp*3*x+5), cl], can[(1 + n_samp*3*x + 5):(n_samp*3*x+n_samp*2), cl],  exact = F)$p.value)
tpv2 = sapply(c(0:9), function(x) wilcox.test(can[(1 + n_samp*3*x):(n_samp*3*x+n_samp), cl], can[(1 + n_samp*3*x + n_samp*2):(n_samp*3*x+n_samp*3), cl],  exact = F)$p.value)
tpv3 = sapply(c(0:9), function(x) wilcox.test(can[(1 + n_samp*3*x + n_samp):(n_samp*3*x+n_samp*2), cl], can[(1 + n_samp*3*x + n_samp*2):(n_samp*3*x+n_samp*3), cl],  exact = F)$p.value)
sum(c(tpv1, tpv2, tpv3) < 0.05)
```

Number of plots that have significant differences within ones according to Kruskal-Wallis test
```{r}
cl =  6
krpv = sapply(c(0:9), function(x) kruskal.test(can[(1 + n_samp*3*x):(n_samp*3*x+n_samp), cl], can[(1 + n_samp*3*x + n_samp):(n_samp*3*x+n_samp*2), cl], can[(1 + n_samp*3*x + n_samp*2):(n_samp*3*x+n_samp*3), cl])$p.value)
sum(krpv < 0.05)
```

## Differences within plots and between plots

Plots show mean of a variable with confidence intervals

```{r, message=FALSE, warning=FALSE}
ynames = colnames(can)[3:ncol(can)]

plot_diff(can, ynames[1])
plot_diff(can, ynames[2])
plot_diff(can, ynames[3])
plot_diff(can, ynames[4])
plot_diff(can, ynames[5])

```
## Dry/wet

Mann-Witney shows significant differences between dry and lake-bed in means of 
```{r}
mwpv = sapply(c(3,5,6), function(x) wilcox.test(can[1:150, x], can[151:300,x],  exact = F)$p.value)
colnames(can)[c(3,5,6)[which(mwpv < 0.05)]]
```

# Individual plant traits


```{r Reading the individual plant traits sheet of plant_traits table, message=FALSE, warning=FALSE}
ind_path <- "/home/lena/Tundra_exp/plant_traits_DPE.xlsx"
ind <- read.xlsx(ind_path, 2)

ind = ind[!is.na(ind$block),]
ind = ind[,-12]
cols = c(6:ncol(ind));    
ind[,cols] = apply(ind[,cols], 2, function(x) as.numeric(as.character(x)))
ind[,1] = factor(ind[,1],levels(ind[,1])[c(1,3:10,2)])


rm(clmn, ind_path)
```

## Differences within plots and between plots

Plots show mean of a variable with confidence intervals

```{r, message=FALSE, warning=FALSE}
ynames = colnames(ind)[cols]

plot_diff(ind, ynames[1])
plot_diff(ind, ynames[2])
plot_diff(ind, ynames[3])
plot_diff(ind, ynames[4])
plot_diff(ind, ynames[5])
plot_diff(ind, ynames[6])
plot_diff(ind, ynames[7])
plot_diff(ind, ynames[8])
plot_diff(ind, ynames[9])
plot_diff(ind, ynames[10])
plot_diff(ind, ynames[11])
plot_diff(ind, ynames[12])
plot_diff(ind, ynames[13])
```

## Dry/wet

Mann-Witney shows significant differences between dry and lake-bed in means of 
```{r}
mwpv = sapply(cols, function(x) wilcox.test(ind[ind$site == "ridge", x], ind[ind$site == "lakebed", x],  exact = F)$p.value)
colnames(ind)[cols[which(mwpv < 0.05)]]
```

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).
